---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Talyah Pierce"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```
---

## The question

Every number the American Community Survey publishes is an estimate drawn from a sample, not a count. That means every number comes with a margin of error — a statement about how wrong it might be. Most people ignore it. In this assignment, you won't.

In this analysis, I am looking at the estimated number of individuals with bachelors degrees in all 14 counties across Massachusetts. Here, I will look at how counties of varying population sizes produce more or less reliable estimates of population characteristics, specifically the number of individuals with bachelors degrees. This analysis will shed light on how the surveys that rely on random samples to generalize population characteristics, like the American Community Survey, may inaccurately represent areas with smaller populations.

---

## Choosing what to measure

```{r}
my_state <- "MA"       
my_variable <- "B15003_022"    
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

```{r}
dim(county_data)
head(county_data)
```

After running the line of code above, the data returned back 14 rows, indicating 14 counties in Massachusetts. This matches the number of counties listed online.

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000 than for an estimate of 200,000. To compare across places, we need it as a percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

The area with the largest margin of error relative to its estimate is Dukes County, with a margin of error percent approximately 17.6%. Because Dukes County has a smaller estimate and higher margin of error, this makes the ACS estimate less reliable.

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```


Most of the counties fall into the high confidence level category, with 10 counties in Massachusetts having a margin of error percent below 5%. Only 2 counties, Berkshire and Franklin, have a moderate confidence level between 5% and 10%. Additionally, only 2 counties, Dukes and Nantucket, have a low confidence level with a margin of error percent above 10%.

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(bachelors = my_variable,
                pop = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = bachelorsM / bachelorsE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```



The average population of high-confidence counties in Massachusetts is 675,837. The average population of low-confidence counties in Massachusetts is 20,751; however, it is worth nothing that only one county is considered in this average. This suggests that counties with higher populations have lower margins of error, producing more reliable estimates. At the county-level, due to the larger size of most counties, this is not a glaring issue in many cases. However, when working with more granular geographies (i.e., tract-level, block group-level), this effect can produce misleading estimates.

---

## What this means for policy

Based on the analysis, the two counties we know the least about are Dukes County and Nantucket County. Interestingly enough, both counties are small island counties with active county governments, home to Nantucket and Martha’s Vineyard.

The American Community Survey (ACS) works by targeting 3.5 million addresses annually as opposed to collecting data from all households in decade intervals like the Census. Because of this, the ACS relies on these samples to produce representative estimates at each geographic level, excluding block groups. The ACS has two survey versions: a one-year survey, which provides estimates for areas with at least 65,000 people, and a five-year survey, which provides estimates of all population-level areas. Because the ACS relies on a random sample of addresses to accurately estimate data points, smaller geographic areas receive less representation, making them more likely to have less representative estimates.

If a state agency were to view data across differently populated areas the same, this would result in less-populated, likely rural areas from receiving fair representation in receiving state resources. For example, if an agency were to rely on ACS estimates to fairly allocate housing funds to Scranton, with a population of around 75,000 people, vs. Wellsboro, with a population of 3,472 people, an area like Wellsboro would like receive more variable/inconsistent funding compared to an area like Scranton with a larger population.

---

## Changing the geographic scale
   
```{r}

# run the same code but at the tract-level for Barnstable County #

county_wide_tract <- get_acs(
  geography = "tract",
  variables = c(bachelors = my_variable,
                pop = "B01003_001"),
  state = my_state,
  county = "Barnstable",
  year = 2023,
  survey = "acs5",
  output = "wide"
)

# compute the margin of error percent for bachelors degrees in Barnstable #

barnstable_moe <- county_wide_tract %>%
  mutate(moe_pct = bachelorsM / bachelorsE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 20  ~ "High confidence",
    moe_pct < 30 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(barnstable_moe, reliability)

head(barnstable_moe)

# compare different levels or reliability in Barnstable #

barnstable_moe %>%
  group_by(reliability) %>%
  summarize(
    n_tracts = n(),
    avg_population = mean(popE)
  )

```


In the code above, I chose to calculate the estimated count of individuals in Barnstable, MA that have a bachelors degree at the tract-level. To calculate the confidence levels, I adjusted the threshold for what is considered high confidence to be slightly more flexible, since tracts are a smaller geographic area than counties and inherently require a bit more flexibility.Thus, tracts with high confidence level have a margin of error percent less than 20%, tracts with a moderate confidence level have a margin of error percent between 20% and 30%, and tracts with a low confidence level have a margin of error percent greater then 30%. After grouping tracts by confidence levels and calculating the average population in each, it is clear that the pattern from the previous exercise holds. Tracts with higher confidence levels have an average population of 4,326 per tract, whereas areas with low confidence levels have an average population of 3,730.

This supplemental analysis provides evidence that discrepancies in population even at a smaller scale can yield even bigger disparities in estimates. While at the county-level a 100 or 1,000 person difference in population may not yield vastly different confidence levels, at the tract-level, this can result in glaring disparities in confidence level estimates.

---
